---
title: "STATS 762 Week 9"
author: " "
date: "May 15, 2019"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preparation

First I loaded required packages and load datasets.

```{r}
library(glmnet); #Ridge and Lasso  
library(grpreg); #Group lasso
ozone=read.csv(file='Ozone35.csv',header=TRUE); 
ozone=as.matrix(ozone); #put in a matrix format - Design matrix should be read as a matrix for glmnet
binary=read.csv(file='binary.csv',header=TRUE);
binary=as.matrix(binary); 
data(Birthwt) #dataset in grpreg
```

# Ridge regression

```{r}
#Slide 13
ridge.lam1=glmnet(ozone[,-1],ozone[,1],alpha=0,lambda=1,standardize=TRUE)
#Slide 15
ridge.fit=glmnet(ozone[,-1],ozone[,1],alpha=0,standardize=TRUE)
plot(ridge.fit)

```

# Lasso regression
```{r}
#Slide 21
lasso.lam1=glmnet(ozone[,-1],ozone[,1],alpha=1,lambda=1,standardize=TRUE)
#Slide 23
lasso.fit=glmnet(ozone[,-1],ozone[,1],alpha=1,standardize=TRUE) 
plot(lasso.fit)
```

#Optimal lambda using CV 
##Ridge regression 
```{r}
#Slide 30 and 31
cv.ridge=cv.glmnet(as.matrix(ozone[,-1]),as.matrix(ozone[,1]),alpha=0,standardize=TRUE)
plot(cv.ridge)
#lambda minimizing the cv error
cv.ridge$lambda.min
#Min cv error
cv.ridge$cvm[cv.ridge$lambda==cv.ridge$lambda.min]
#lambda 1sd - cv error is off by 1sd of the min error
cv.ridge$lambda.1se
#cv error off by 1sd of the min er
cv.ridge$cvm[cv.ridge$lambda==cv.ridge$lambda.1se]
#Coefficients when lambda.1se
coef(cv.ridge,s=cv.ridge$lambda.1se)
```
##Lasso regression 
```{r}
#Slide 30 and 31
cv.lasso=cv.glmnet(as.matrix(ozone[,-1]),as.matrix(ozone[,1]),alpha=1,standardize=TRUE)
plot(cv.lasso)
#lambda minimizing the cv error
cv.lasso$lambda.min
#Min cv error
cv.lasso$cvm[cv.lasso$lambda==cv.lasso$lambda.min]
#lambda 1sd - cv error is off by 1sd of the min error
cv.lasso$lambda.1se
#cv error off by 1sd of the min er
cv.lasso$cvm[cv.lasso$lambda==cv.lasso$lambda.1se]
#Coefficients when lambda.1se
coef(cv.lasso,s=cv.lasso$lambda.1se)
```

#Elastic net 
```{r}
#Slides 35 and 36
#alpha=0.2
elastic.fit=glmnet(ozone[,-1],ozone[,1],alpha=0.2,standardize=TRUE)
plot(elastic.fit)
cv.elastic=cv.glmnet(as.matrix(ozone[,-1]),as.matrix(ozone[,1]),alpha=0.2,standardize=TRUE)
plot(cv.elastic)
```

#Logistic reg - Graduate school data
```{r}
#Slide 42
binary.fit=glmnet(binary[,-1],binary[,1],family='binomial',alpha=1,standardize=TRUE)
plot(binary.fit)
binary.lasso=cv.glmnet(binary[,-1],binary[,1],family='binomial',alpha=1,standardize=TRUE)
plot(binary.lasso)
#optimal lambda
binary.lasso$lambda.1se
#coefficients when lambda.1se
coef(binary.lasso,s=binary.lasso$lambda.1se)

```

#Group lasso
```{r}
#Slide 51
Birthwt.fit <- grpreg(Birthwt$X, Birthwt$bwt, Birthwt$group, family='gaussian',penalty="grLasso")
plot(Birthwt.fit)

#Slide 53
#prediction
predict(Birthwt.fit,Birthwt$X,type='response',lambda=0.1)[1:10]
predict(Birthwt.fit,Birthwt$X,type='group',lambda=0.1)
predict(Birthwt.fit,Birthwt$X,type='coefficients',lambda=0.1)[1:10]

#Slide 54
plot(Birthwt$bwt,predict(Birthwt.fit,Birthwt$X,type='response',lambda=0.1),xlim=c(0.5,5.5),ylim=c(0.5,5.5))

#Slide 56
cvfit <- cv.grpreg(Birthwt$X, Birthwt$bwt, Birthwt$group, family='gaussian', penalty="grLasso")
coef(cvfit,s=cvfit$lambda.1se)

#Slide 57
plot(Birthwt$bwt,predict(cvfit,Birthwt$X,type='response',lambda=cvfit$lambda.min)
                         ,xlim=c(0.5,5.5),ylim=c(0.5,5.5))

```


#Exercise 

Data (evap.txt) was recorded for factors affecting the evaporation water
from a location in Texas. The first column "evap"" is a response variable. See Arden's Lecture slides 7 for details. 

```{r}
evap=read.delim('evap.txt',sep="",header=TRUE)
```

(a) Find the optimal subset of covariates to predict the amount of evaporation from soil.
```{r}
```
(b) Find the coefficients for your model in (a).
```{r}
```
(c) Plot the actual data against prediction using your model in (a).
```{r}
```

